\section*{Application I: Human 1000 Genomes Novel Analyses}

We applied the full GraphPop analytical stack to the 1000 Genomes Project Phase~3 dataset\cite{1000genomes2015global}: 3{,}202 samples across 26 sub-populations, 22 autosomes, $\sim$70.7M Variant nodes, 91{,}973 Gene nodes, 1.46M GenomicWindow nodes, and 2{,}241 Reactome pathways\cite{gillespie2022reactome} annotated via \texttt{HAS\_CONSEQUENCE} and \texttt{IN\_PATHWAY} edges. The scale and annotation depth of this graph enable a mode of investigation inaccessible to tool-based pipelines: queries that traverse from individual genotypes through variant consequences through gene pathways through population-level selection statistics in a single step.

\subsection*{Pathway-level population differentiation}

Population differentiation is conventionally computed at the variant or gene level; pathway-level Fst requires aggregating per-variant estimates across gene sets, a multi-step join across at least three tools (VCFtools, bedtools, pathway database API). In GraphPop, the equivalent is a single traversal:

\begin{center}
\texttt{MATCH (p:Pathway)$\leftarrow$[:IN\_PATHWAY]-(g:Gene)$\leftarrow$[:HAS\_CONSEQUENCE]-(v:Variant)}\\
\texttt{RETURN p.name, mean(v.fst\_yri\_ceu) ORDER BY 2 DESC}
\end{center}

Ranking 2{,}241 pathways by mean Hudson Fst (YRI vs CEU) placed known pigmentation biology at the top: the SLC24A5-linked oculocutaneous albinism pathway (mean Fst~$= 0.452$), consistent with strong post-OoA selection on skin pigmentation in European populations\cite{yi2010sequencing}. Cardiac ion channel pathways formed a functionally coherent cluster in the top quartile: TREK/TWIK potassium channels (Fst~$= 0.162$), Phase 3 rapid repolarisation (Fst~$= 0.133$), and voltage-gated potassium channels---implicating KCNE1, KCNE4, KCNH2, and KCNQ1 as population-differentiated cardiac loci (Supplementary Fig.~\ref{sfig:pathway_fst}).

This pathway-level view is impossible without annotation conditioning: the KCNE1 locus, for instance, contributes to cardiac repolarisation but is co-localised with variants classified as metabolic and immune pathway members. Only by querying pathway membership directly in the graph can one separate these signals without manual curation.

\subsection*{Graph-native detection of convergent positive selection}

XP-EHH scores for 325 population pairs ($\times$22~chromosomes) are stored as properties on Variant nodes. Querying for genes under sweep in multiple independent continental lineages---convergent positive selection---reduces to a single Cypher pattern match on stored node properties, requiring no re-computation or file intersection. We identified \textbf{9 genes with Garud $H_{12} > 0.3$ in $\geq$2 independent continental groups} (Fig.~\ref{fig:human_selection}).

The most striking signal was \textbf{KCNE1}, the only gene showing sweep evidence in all five continental groups ($H_{12} = 1.0$ in some populations; Fst ratio between- vs within-group $= 0.296$). The elevated between-group Fst combined with universally high $H_{12}$ is the signature of an ancient, pre-OoA selective sweep that reached near-fixation before the Out-of-Africa bottleneck\cite{sabeti2007genome}---a signal recoverable only by cross-querying independently computed statistics stored on the same graph. The classical equivalent requires intersecting five separate XP-EHH output files, then joining against Garud's $H$ scan results and Fst values via chromosome coordinate matching.

Consequence-level analysis reinforced this picture: HIGH-impact variants (missense, stop-gain) showed significantly lower mean Fst (0.124) than LOW-impact variants (0.143; Mann--Whitney $U$, $p \approx 0$), confirming genome-wide purifying selection on functionally deleterious alleles---a population-level signal accessible in a single \texttt{HAS\_CONSEQUENCE}-conditioned query.

\subsection*{Individual-level evolutionary trajectories}

Standard population genetics PCA (PLINK, EIGENSOFT) projects individuals onto axes of ancestry derived from allele frequency differences. We introduce a complementary representation: each individual is characterised by a vector of five evolutionary process statistics (genome-wide FROH, total ROH count, chromosome~22 heterozygosity rate, chr22 rare variant burden, and chr22 rare missense burden), then embedded via PCA and UMAP. This \emph{individual process space} captures the evolutionary forces acting on each individual rather than their ancestral coordinates.

Applied to all 3{,}202 samples (Fig.~\ref{fig:individual}), the first two statistical PCs explained 57.0\% and 30.9\% of variance respectively. \textbf{PC1 is a diversity axis}: African samples, with the largest long-term $N_e$, score highest on heterozygosity and rare burden (mean rare missense per individual: African $= 34.5$, European $= 16.8$, East Asian $= 18.3$). \textbf{PC2 is an inbreeding axis}: South Asian samples score highest on FROH (mean $= 0.014$), consistent with documented consanguinity practices. Individual positions strongly predict their population's position in population-level process space ($\rho = 0.74$, Spearman), validating the approach.

Outlier analysis revealed HG01816 (CDX, Chinese Dai) as the most extreme individual in process space (z-distance $= 9.50$ from population centroid), carrying genome-wide FROH~$= 0.125$---18$\times$ the CDX population mean---consistent with recent close-relative mating (Supplementary Fig.~\ref{sfig:outlier}). GraphPop enables immediate follow-up: a single 3-hop Cypher query from this sample through CARRIES edges to HAS\_CONSEQUENCE edges identifies the specific rare variants driving its anomalous burden, with gene and pathway annotation, in milliseconds.

\subsection*{Bottleneck-driven purging is pathway-specific}

The African populations carry $\sim$2$\times$ more rare missense variants per individual than non-African populations, reflecting larger long-term effective population size. A key question---not addressable by standard tools---is whether this enrichment is uniform across all biological functions or concentrated in specific pathway classes. Answering it requires simultaneously accessing individual genotypes (CARRIES edges), variant consequences (HAS\_CONSEQUENCE), pathway membership (IN\_PATHWAY), and population-level pathway differentiation statistics (from prior Fst computations): four layers of information co-located only in the graph.

We classified 5{,}833 chr22 missense variant positions as belonging to either \emph{constrained pathways} (lowest-quartile mean Fst; maintained by purifying selection across all populations) or \emph{divergent pathways} (highest-quartile Fst; targets of population-specific selection). For each of the 3{,}202 individuals, we computed the constrained:divergent (C:D) burden ratio from haplotype arrays.

The C:D ratio was 2.2$\times$ higher in African samples (0.187) than in East Asian samples (0.086), and decreased monotonically across the bottleneck gradient: $\rho(\text{FROH},\, \text{C:D ratio}) = -0.065$ ($p = 2.1 \times 10^{-4}$; Fig.~\ref{fig:purifying}). This pattern---bottlenecked populations disproportionately depleted of constrained-pathway rare missense---is consistent with drift acting like purifying selection in small populations, but preferentially at functionally critical loci: variants in essential gene sets face stronger selection against them, so drift removes them more efficiently when $N_e$ falls. The gradient is detectable only through individual-level, pathway-stratified burden analysis, which no existing tool pipeline provides as a single operation.

\subsection*{Novel analyses enabled: a capability comparison}

Table~\ref{tab:novel} summarises the analyses demonstrated above alongside the classical equivalent workflow. In each case, the graph-native approach replaces a multi-tool pipeline with a single traversal or a combination of pre-stored results. Three of the six analyses have no classical equivalent at all---they ask questions that are practically infeasible without co-location of individual genotypes, variant annotations, and population-level selection statistics.

\begin{table}[h]
\centering
\small
\caption{Novel analyses enabled by GraphPop's co-located individual genotype, annotation, and selection graph.}
\label{tab:novel}
\begin{tabular}{p{4.2cm} p{4.8cm} p{4.8cm}}
\toprule
\textbf{Analysis} & \textbf{Classical approach} & \textbf{GraphPop} \\
\midrule
Pathway-level population Fst & VCFtools Fst $\to$ BEDtools gene overlap $\to$ pathway API $\to$ manual join (3 tools + join) & Single \texttt{MATCH} traversal through \texttt{IN\_PATHWAY} \\
\addlinespace
Convergent sweep detection & 5 XP-EHH runs $\to$ 5 output files $\to$ file intersection $\to$ coordinate join & Cypher pattern match on stored \texttt{xpehh\_*} node properties \\
\addlinespace
Individual process-space embedding & PLINK ROH $\to$ custom het script $\to$ VEP $\to$ rare variant filter $\to$ manual join & Pre-stored FROH + single haplotype array pass \\
\addlinespace
Outlier $\to$ driver variant trace & Extract single-sample VCF $\to$ VEP $\to$ pathway join & 3-hop Cypher from Sample $\to$ CARRIES $\to$ HAS\_CONSEQUENCE $\to$ IN\_PATHWAY \\
\addlinespace
Temporal burden vector (ancient vs recent sweep) & Not feasible (requires iHS-by-timescale classification + per-individual variant access across full genome) & iHS stored on Variant nodes; CARRIES edges give per-individual access \\
\addlinespace
Pathway-stratified purging gradient & Not feasible (requires simultaneous genotype $\times$ consequence $\times$ pathway $\times$ FST access) & Single session: haplotype arrays $+$ \texttt{IN\_PATHWAY} query $+$ prior Fst results \\
\bottomrule
\end{tabular}
\end{table}

\subsection*{Biological synthesis}

The 1000 Genomes analysis yields three conclusions that are specific to graph-native investigation:

\begin{enumerate}
\item \textbf{KCNE1 underwent a universal pre-OoA sweep.} Convergent $H_{12}$ peaks in all five continental groups, combined with an elevated between-group:within-group Fst ratio, identify KCNE1 as an ancient sweep predating the Out-of-Africa bottleneck---a finding requiring cross-query of independently computed statistics stored as node properties. No VCF-centric pipeline produces this inference without manual file merging.

\item \textbf{Individual evolutionary trajectories reveal intra-population heterogeneity invisible to PCA.} The 5-dimensional process space (FROH, ROH count, het rate, rare burden, rare missense) separates individuals by evolutionary force rather than ancestry, identifies extreme inbreeding outliers ($\text{FROH} = 18 \times$ population mean), and directly implicates the variants driving each individual's anomalous position through graph traversal.

\item \textbf{Bottleneck-driven purging is pathway-specific.} The 2.2$\times$ excess of constrained-pathway rare missense in African vs East Asian individuals, and its genome-wide FROH correlation ($\rho = -0.065$, $p = 2.1 \times 10^{-4}$), demonstrate that demographic drift preferentially depletes rare functional variants from conserved biological systems. This claim required the simultaneous availability of individual genotypes, variant consequences, pathway annotations, and population Fst statistics---a data co-location achievable only through the graph.
\end{enumerate}

\section*{Application II: Rice 3K Genome Analysis}

We applied all 12 GraphPop procedures to the 3K Rice Genomes Project\cite{wang2018genomic} dataset (29.6M SNPs, 3{,}024 accessions, 12 subpopulations across 12 chromosomes). The graph database (46~GB) was constructed in a single import pipeline run. All downstream analyses were pure Cypher procedure calls---no file export, no format conversion, no external tool invocation. The analysis demonstrates that graph-native computation does not merely reproduce known biology faster, but reveals evolutionary dynamics that have not been systematically characterized because no existing tool makes the necessary analyses practical.

\subsection*{Correctness validation}

Standard statistics reproduced the established biology of Asian rice. The UPGMA tree from pairwise W\&C Fst recovered the indica--japonica split with Fst~$= 0.71$ between GJ-tmp and XI-1A (the most divergent pair; Fig.~\ref{fig:rice}c). Indica (XI) subpopulations were 2--3$\times$ more diverse than japonica (GJ), with GJ-tmp the least diverse ($\pi = 0.019$, Tajima's $D = -2.41$, $F_{IS} = 0.72$, FROH~$= 0.080$)---matching the known temperate japonica bottleneck history. XP-EHH and Garud's $H$ recovered known domestication genes (GW5, Hd1, PROG1, Wx, OsC1, DRO1) at expected genomic positions. These results confirm numerical accuracy (validated to $< 0.000001\%$ error against scikit-allel on 1000 Genomes chr22) and establish that GraphPop operates correctly on a full-scale non-model genome.

\subsection*{First systematic genome-wide quantification of the cost of domestication}

The ``cost of domestication''---accumulation of mildly deleterious mutations due to population bottlenecks---has been demonstrated in rice through interspecific comparisons with wild relatives\cite{lu2006cost}. However, to our knowledge, $\pi_N/\pi_S$ has not been systematically quantified across all subpopulations of cultivated rice at genome scale, because doing so requires annotation-conditioned diversity computation that no existing tool provides as a single operation.

By calling \texttt{graphpop.diversity} with \texttt{\{consequence: `missense\_variant'\}} and \texttt{`synonymous\_variant'} for all 12 populations $\times$ 12 chromosomes (288 procedure calls), we obtained $\pi_N/\pi_S$ ratios for every subpopulation (Fig.~\ref{fig:rice}a). \textbf{All 12 subpopulations show $\pi_N/\pi_S > 1.0$} (range: 1.018--1.146; Supplementary Table~\ref{tab:rice_pinpis}), demonstrating that genome-wide relaxation of purifying selection is universal across cultivated Asian rice. The pattern is biologically informative:

\begin{itemize}
\item Admixed groups show the highest ratios (admix: 1.146, GJ-adm: 1.145), consistent with accumulation of slightly deleterious alleles from multiple bottlenecked source populations.
\item GJ-tmp (temperate japonica) shows the lowest ratio (1.018) despite having the strongest demographic bottleneck ($\pi = 0.019$, FROH = 0.080), suggesting that intense directional selection during temperate adaptation has partially counteracted the relaxation of purifying selection.
\item Indica subpopulations cluster narrowly (1.053--1.079), reflecting their larger effective population size where purifying selection remains more efficient.
\end{itemize}

This population-level resolution of the cost of domestication---distinguishing admixed load accumulation from selection-constrained bottleneck survivors---is a direct product of annotation-conditioned computation. The classical equivalent would require 288 multi-step pipeline runs (VEP $\to$ filter $\to$ subset VCF $\to$ compute $\pi$), making systematic analysis impractical.

\subsection*{First genome-wide detection of adaptive protein divergence between rice ecotypes}

Whether protein-coding divergence between rice subpopulations exceeds neutral expectation---evidence that natural or artificial selection has driven amino acid changes---has been examined at candidate loci but not systematically at genome scale. Annotation-conditioned genome scans make this investigation straightforward.

By running \texttt{graphpop.genome\_scan} with \texttt{\{consequence: `missense\_variant'\}} and \texttt{`synonymous\_variant'} separately for three major population pairs, we produced whole-genome Fst profiles restricted to each variant class (Fig.~\ref{fig:conditioned}a; Fig.~\ref{fig:rice}b). All three pairs showed missense Fst exceeding synonymous Fst:

\begin{itemize}
\item GJ-tmp vs GJ-trp: ratio 1.106 (the temperate--tropical japonica split, $\sim$4.3--5.5~kya\cite{alam2025genomics})
\item GJ-tmp vs XI-1A: ratio 1.044 (the japonica--indica split)
\item XI-1A vs cA-Aus: ratio 1.028 (divergence within the indica/aus complex)
\end{itemize}

The strongest signal at the temperate--tropical japonica boundary indicates that rapid ecological specialization drove adaptive protein evolution---a finding consistent with the phenotypic divergence between these ecotypes (cold tolerance, photoperiod sensitivity, grain quality) but not previously demonstrated at the genome-wide level via direct missense/synonymous Fst comparison.

The top missense-Fst outlier windows (e.g., Chr1:7.45~Mb, Fst~$= 0.966$ for XI-1A vs cA-Aus) pinpoint candidate loci for adaptive protein evolution, invisible to any tool lacking annotation conditioning.

\subsection*{Multi-statistic convergence analysis via the persistent analytical record}

Because all computed results reside on the same graph as permanent node properties, GraphPop enables a mode of analysis impossible with ephemeral outputs: querying for convergent signals across independently computed statistics, without re-computation or file merging.

After running diversity, divergence, genome scan, XP-EHH, PBS, Garud's $H$, nSL, Fay \& Wu's $H$, and ROH across all populations, we queried the graph for loci where multiple statistics converged (Fig.~\ref{fig:selection}c). At \textbf{Chr6:9.0--9.6~Mb} (Hd1, heading date\cite{huang2012map}), the same Variant nodes carry: XP-EHH~$= -4.15$ (GJ-trp vs XI-1A), PBS~$= 1.10$ (GJ-trp), and Fay \& Wu's $H = -0.492$ in a nearby window---three independent lines of evidence for selection, all retrievable from already-stored properties. At \textbf{Chr5:5.3~Mb} (GW5, grain width), XP-EHH peaks in two independent comparisons (XI-1A vs cA-Aus: 4.91; GJ-trp vs XI-1A: $-3.56$) cross-validate each other through the same graph query.

Graph traversal connects these statistics directly to genes: known domestication loci (GW5, Hd1, PROG1, Wx, OsC1, DRO1\cite{huang2012map,alam2025genomics}) were recovered by following \texttt{HAS\_CONSEQUENCE} edges from peak Variant nodes to Gene nodes---zero additional tools, zero coordinate matching (Table~\ref{tab:genes}; Fig.~\ref{fig:selection}a,b).

In classical pipelines, this three-way convergence analysis would require merging outputs from scikit-allel (EHH), a custom implementation (Garud's $H$), VCFtools (Fst), and bedtools (annotation) via coordinate joins. In GraphPop, it is a Cypher query.

\subsection*{Biological synthesis}

The rice 3K analysis reveals three layers of evolutionary insight enabled by graph-native investigation:

\begin{enumerate}
\item \textbf{The cost of domestication is universal but unequal.} All 12 subpopulations show $\pi_N/\pi_S > 1.0$, but the pattern distinguishes admixed load accumulation from selection-constrained bottleneck survivors---a resolution achievable only through systematic annotation-conditioned analysis.

\item \textbf{Adaptive protein evolution accompanies ecological divergence.} Missense Fst exceeds synonymous Fst at all major population boundaries, with the strongest signal at the temperate--tropical split---evidence for directional selection on protein-coding variants during recent ecological specialization, detectable only via annotation-conditioned genome scans.

\item \textbf{Multi-statistic convergence validates and strengthens selection candidates.} Domestication genes appear as peaks in independently computed statistics (XP-EHH, PBS, Fay \& Wu's $H$, Garud's $H_{12}$), cross-referenced to gene annotations---all discoverable by querying the persistent graph record.
\end{enumerate}

The entire analysis---from import through all 12 procedure types to biological interpretation---required no tool outside GraphPop and Neo4j. Results remain in the graph, publicly available for any future investigation.
