\documentclass[10pt]{article}

% Nature Methods formatting
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{lineno}
\usepackage{authblk}
\usepackage{siunitx}
\usepackage{multirow}
\usepackage{array}
\usepackage{longtable}
\usepackage{microtype}

\linenumbers

% Title
\title{\textbf{GraphPop: graph-native population genomics with annotation-conditioned queries}}

\author[1]{Jianfeng Mao}
% Add co-authors as needed
% \author[2]{Author Two}

\affil[1]{Affiliation}
% \affil[2]{Affiliation Two}

\date{}

\begin{document}

\maketitle

% Abstract
\input{abstract}

% Main text
\input{introduction}
\input{architecture}
\input{benchmarks}
\input{application}
\input{discussion}

% Methods
\section*{Online Methods}
\input{methods}

% Data and Code Availability
\section*{Data and Code Availability}
GraphPop source code, including all 12 stored procedures, the import pipeline, the MCP server, and the ToolUniverse integration, is available at \url{https://github.com/[repository]} under the MIT license.

Pre-built Neo4j graph databases for both datasets analyzed in this study are publicly available for immediate use: the 1000 Genomes Project full-genome graph (chr1--22, 3{,}202 samples) and the 3K Rice Genomes Project graph (12 chromosomes, 3{,}024 accessions, 29.6M variants) with all computed statistics (diversity, divergence, selection scans, ROH, genome scan windows) persisted as node properties. These databases are deposited at Figshare/Zenodo [DOI to be assigned upon acceptance]. Researchers can download and load these graphs directly into a Neo4j instance to query any pre-computed result or run additional GraphPop analyses without re-importing from VCF.

The source VCF data are available from the 1000 Genomes Project (\url{https://www.internationalgenome.org/}) and the 3K Rice Genomes Project (\url{https://snp-seek.irri.org/}; Wang et al.\ 2018). Reactome pathway annotations are from \url{https://reactome.org/}. Ancestral allele annotations were derived from Ensembl EPO multi-species alignment FASTA files. All benchmark scripts, analysis scripts, and figure-generation code are included in the repository.

% Acknowledgements
\section*{Acknowledgements}
We thank the 1000 Genomes Project Consortium and the 3K Rice Genomes Project for making their data publicly available.

% Author Contributions
\section*{Author Contributions}
J.M.\ conceived the project, designed the architecture, implemented all software, performed analyses, and wrote the manuscript.

% Competing Interests
\section*{Competing Interests}
The authors declare no competing interests.

% References
\bibliographystyle{naturemag}
\bibliography{references}

\clearpage

% Supplementary
\input{supplementary}

\end{document}
